### 2017-07-07 opsem.tex
Opsem txnid added to local reduction.
Edits.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). The $\I$ in the \C{TXN} block syntax is the
transaction's isolation specification, which will be explained later.
Certain terms that only appear at run-time are also present in $c$.
These include a {\sf txn} block tagged with sets ($\stl$ and $\stg$) of
records whose meaning is explained below, and a runtime {\sf foreach}
expression that keeps track of the set ($s_1$) of items already
iterated, and the set ($s_2$) of items yet to be iterated. Note that
even the surface-level syntax of the \C{FOREACH} command shown here is
a little different from the one used in previous sections; its
higher-order function has two arguments, $y$ and $z$, which are
invoked (during the reduction) with the set of already-iterated items,
and the current item, respectively. This form of \C{FOREACH} lends
itself to inductive reasoning that will be useful for verification
(Sec.~\ref{sec:reasoning}).

@@

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward.  Records in $\stg$ are uniquely identifiable
through their $\idf$ field, which is auto-generated and does not
belong to the surface language, i.e., $\C{id}\notin f$. For a set $S$
of records, we define $\dom(S)$ as the set of unique ids of all 
records in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, databases, in reality, enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, which are identified by a hidden $\delf$ field set to
\C{true}. When the transaction commits, its local database is
atomically \emph{flushed} to the global database, committing these
heretofore uncommitted records. The flush operation ($\gg$) is defined as
follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, but the update is not
a deletion ($\neg r.\delf$). Thus, flush defines the result of
atomically applying a transaction's local writes to the global
database.  Besides the commit, flush also helps a transaction read its
own writes. Intuitively, the result of a read operation inside a
transaction must be computed on the database resulting from flushing
the current local state ($\stl$) to the global state ($\stg$). The
abstract machine of Fig.~\ref{fig:txnimp}, however, does not let a
transaction read its own writes. This simplifies the semantics,
without losing any generality, since substituting $\stl\gg\stg$ for
$\stg$ at select places in the reduction rules effectively allows
reads of uncommitted transaction writes to be realized.

The small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. The transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$. The $\I$ in the \C{TXN} block syntax is the
transaction's isolation specification, which will be explained later.
Certain terms that only appear at run-time are also present in $c$.
These include a {\sf txn} block tagged with sets ($\stl$ and $\stg$) of
records whose meaning is explained below, and a runtime {\sf foreach}
expression that keeps track of the set ($s_1$) of items already
iterated, and the set ($s_2$) of items yet to be iterated. Note that
even the surface-level syntax of the \C{FOREACH} command shown here is
a little different from the one used in previous sections; its
higher-order function has two arguments, $y$ and $z$, which are
invoked (during the reduction) with the set of already-iterated items,
and the current item, respectively. This form of \C{FOREACH} lends
itself to inductive reasoning that will be useful for verification
(Sec.~\ref{sec:reasoning}).

@@

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local ($\stl$), or global ($\stg$) database, both
of which are modeled as a set of records of a pre-defined type, i.e.,
they all belong to a single table.  The generalization to multiple
tables is straightforward. The semantics assumes that the records in
$\stg$ are uniquely identifiable through their $\idf$ field, and
enforces this invariant wherever necessary. Certain hidden fields are
treated specially by the operational semantics, and they are hidden
from the surface language. These include a $\txnf$ field that tracks
the identifier of the transaction that last updated the record, and a
$\delf$ field that flags the deleted records in $\stl$.  For a set $S$
of records, we define $\dom(S)$ as the set of unique ids of all
records in $S$. Thus $|\dom(\stg)| = |\stg|$. During its execution, a
transaction may write to multiple records in $\stg$. Atomicity
dictates that such writes should not be visible in $\stg$ until the
transaction commits. We therefore associate each transaction with a
local database ($\stl$) that stores such uncommitted
records\footnote{While SQL's \C{UPDATE} admits writes at the
  granularity of record fields, most popular databases enforce
  record-level locking, allowing us to think of ``uncommitted writes''
  as ``uncommitted records''. }. Uncommitted records include deleted
records, whose $\delf$ field is set to \C{true}. When the transaction
commits, its local database is atomically \emph{flushed} to the global
database, committing these heretofore uncommitted records. The flush
operation ($\gg$) is defined as follows:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and has not been updated in $\stl$, i.e., $r.\idf
\notin \dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new
record, or an updated version of an old record, but the update is not
a deletion ($\neg r.\delf$). Thus, flush defines the result of
atomically applying a transaction's local writes to the global
database.  Besides the commit, flush also helps a transaction read its
own writes. Intuitively, the result of a read operation inside a
transaction must be computed on the database resulting from flushing
the current local state ($\stl$) to the global state ($\stg$). The
abstract machine of Fig.~\ref{fig:txnimp}, however, does not let a
transaction read its own writes. This simplifies the semantics,
without losing any generality, since substituting $\stl\gg\stg$ for
$\stg$ at select places in the reduction rules effectively allows
reads of uncommitted transaction writes to be realized.

The small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. The transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly. \C{INSERT} adds a new record
to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
the records in $\stg$ that match the search criteria defined by its
boolean function argument, and adds the records to $\stl$ after
marking them for deletion. \C{SELECT} bounds the name introduced by
\C{LET} to the set of records from $\stg$ that match the search
criteria, and then executes the bound command $c$. \C{UPDATE} uses its
first function argument to compute the updated version of the records
that match the search criteria defined by its second function
argument. Updated records are added to $\stl$.
